---
title: "R programming"
author: "Vinish Shrestha"
date: "10/18/2020"
output:
  beamer_presentation: default
  ioslides_presentation: default
  slidy_presentation: default
---
## 
```{r, echo = FALSE}
x <- "Version0\nThis\tis intro.\nWe will keep building."
writeLines(x) # see also ?strwra
```


## R programming install (ubuntu)
1. install dependencies necessary for adding new repository 
    - sudo apt install dirmngr gnupg apt-transport-https ca-certificates software-properties-common
2. Add Comprehensive R Archive Network (CRAN) repository to the system source list
    - sudo apt-key adv --keyserver keyserver.ubuntu.com --recv-keys E298A3A825C0D65DFD57CBB651716619E084DAB9
    - sudo add-apt-repository 'deb https://cloud.r-project.org/bin/linux/ubuntu focal-cran40/'
3. install R
    - sudo apt install r-base
4. R --version

## Update R 
- $ sudo apt-key adv --keyserver keyserver.ubuntu.com --recv-keys E298A3A825C0D65DFD57CBB651716619E084DAB9
- $ sudo add-apt-repository 'deb https://cloud.r-project.org/bin/linux/ubuntu focal-cran40/'
- $ sudo apt install r-base

Install R studio

- sudo apt-get install gdebi-core
- wget https://download2.rstudio.org/server/bionic/amd64/rstudio-server-1.4.1717-amd64.deb
- sudo gdebi rstudio-server-1.4.1717-amd64.deb


## Install for windows
- visit this [site](https://cran.r-project.org/)
- click download R for windows 
- click on base 
- Then click "Download R 4.1.1 for Windows

Next, you'd wanna install RStudio IDE, for better environment

- go to rstudio [page](https://www.rstudio.com/products/rstudio/download/#download)
- pick the appropriate OS and download
- The current file for windows is RStudio-1.4.1717.exe

Use this [\color{blue}{link}](https://www.datacamp.com/community/tutorials/installing-R-windows-mac-ubuntu) as a helpful resource
Watch this [\color{blue}{video}](https://www.youtube.com/watch?v=NZxSA80lF1I) as a guide if you have problems redo.

## Helpful resources to learn
- [\color{blue}{Venables and friends}](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf)
- [\color{blue}{Grolemund}](https://rstudio-education.github.io/hopr/)
- stackover flow
- CRAN [\color{blue}{https://CRAN.R-project.org}](https://CRAN.R-project.org) 

## 1. Numbers and vectors 
```{r, warning = FALSE, message=FALSE}
x <- c(7.4, 5.2, 10.2, 3.3)
print(x)
```
- x is a variable (object), c() is a function that can store vector arguments
- <- tells that object c receives the designated values; can be replaced by "=" in most instances

```{r, warning = FALSE, message=FALSE}
length(x)
y <- 2*x - 4 #basic arithmetic
print(y)
```

## 2. Regular sequences
```{r, warning=F, message=F}
even_n <- seq(2, 20, 2)
odd_n <- seq(1,19, 2)
even_n
odd_n
```
Repetition

```{r, warning=F, message=F}
rep1 <- rep(1:3, 2)
rep2 <- rep(1:3, each = 2)
rep1
rep2
```

## 3. Missing values
- Often values of variables in observational data are missing 
- missing data is represented by NA
```{r, warning=F, message=F}
x <- c(seq(2, 6, 2), NA)
x
```

Say now you want to drop the missing observation from the vector x and store it in vector y 

```{r, warning=F, message=F}
y <- x[!is.na(x)]
y
```

## 4. Location of a vector 
- use $[i]$, where $i$ is the integer of interest to find out $i^{th}$ element of a vector 

```{r, warning=F, message=F}
even_n <- seq(2, 20, 2)
paste("The second element of even_n is", even_n[2], 
      sep = " ")
```
- here paste joins the character or string "The second element of even_n is" with the object even_n[2]

## Quotes
```{r}
x <- "long\tlines can be\nbroken with newlines"
print(x)
writeLines(x) # see also ?strwra
```
- see ?Quotes 

## 5. R Objects and Class
1. Character vector 
- entered by "

```{r}
a <- c("A", "B","C")
a
class(a)
is.vector(a) #tests whether a is vector
length(a) #returns length
```

## 5. R Objects and Class
2. Numeric 
```{r}
x <- seq(1, 10, 1)
class(x)
length(x) #know the length
x[1] #first element of vector x
```

## 5. R Objects and Class
3. Logicals 

```{r}

a <- 4
l <- a>3
class(l)
```

## 5. R Objects and Class
4. Matrix  

```{r}

mat <- matrix(0, nrow = 3, ncol = 3) #three by three matrix
mat #display the matrix
mat[,1] #display the first column of the matrix
```

## 5.
5. arrays
```{r}
ar <- array(c(11:14, 21:24, 31:34), dim = c(2, 2, 3))
ar
```

## Other objects 
- matrices: multi-dimensional vectors
- factors: efficient way to handle categorical variables
- lists: general form of vector that handles objects of different classes
- data.frames: typical data as in excel 
- functions: used to manipulate data

## A Simple Plot
```{r, fig.align= "center", fig.height = 3, fig.width= 3}
x <- seq(1, 10, 1)
y <- 2 + x*7
plot(x, y)
lines(x, y)
```

##
```{r, echo = FALSE}
nextc <- "End Sept 8; Next time\n 1. dataframe\n 2. manipulating dataframe\n 3. running simple regression "
writeLines(nextc)
```

## 
```{r, echo = FALSE}
z <- " practice\n practice\n practice"
writeLines(z)
```
